---
title: "OffAqua"
subtitle: "HAB monitoring database"
author: "Tim Szewczyk"
organization: "SAMS"
output: 
  html_document:
    theme: spacelab
    df_print: paged
    anchor_sections: TRUE
    toc: yes
    toc_depth: 2
    toc_float: true
  pdf_document:
    toc: yes
  html_notebook:
    theme: spacelab
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=F}
# set directory for knitr as main project directory
knitr::opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
```

```{r setup_workspace, include=FALSE, message=FALSE, warning=FALSE}
# libraries, functions, directories
pkgs <- c("tidyverse", "lubridate", "jsonlite", "sf")
suppressMessages(invisible(lapply(pkgs, library, character.only=T)))
theme_set(theme_bw() + theme(panel.grid.minor=element_blank()))
walk(dir("code", "00[0-9]_fn", full.names=T), source)

gis.dir <- "..\\..\\00_gis\\"
UK_bbox <- st_bbox(c(xmin=-11, xmax=2, ymin=49, ymax=61), crs=st_crs(4326))
coast <- read_sf(paste0(gis.dir, "coastline\\ne_10m_land.shp")) %>%
  add_row(read_sf(paste0(gis.dir, "coastline\\ne_10m_minor_islands.shp"))) %>%
  st_crop(UK_bbox)
```


# Overview

This file is for figuring out how to access the HAB monitoring database. From Steve Gontarek:

> I can give you access to the online DB via these urls – it’s a REST API:
>
> [varro:3001/cefas_counts](varro:3001/cefas_counts)
> [varro:3001/fsa_counts](varro:3001/fsa_counts)
> 
> you can also query them better
>
> [varro:3001/cefas_counts?date_collected=gte.2022-01-01](varro:3001/cefas_counts?date_collected=gte.2022-01-01)
> [varro:3001/cefas_counts?date_collected=gte.2022-01-01](varro:3001/cefas_counts?date_collected=gte.2022-01-01&date_collected=lt.2022-01-18)
>
> e.g. gte ( greater than or equal), lt, less then etc
> 
> more details here for the REST API [Tables and Views — PostgREST 9.0.0 documentation](https://postgrest.org/en/v9.0/api.html)
>
> ...main tables are fsa_counts (Phytoplankton) and cefas_counts (toxins)

Note that access is restricted to the internal SAMS network, so it requires being connected via VPN. I should be able to access what I need using the package `jsonlite`.

I'm not sure how to get R to recognize this form of a url... I've tried a couple of packages and methods, adding file://, etc, but I'm just not sure what I'm doing really. I think varro is the name of the server. So in the meantime, I just copied the current versions to my hard drive using a browser.

```{r full_DB}
fsa.df <- fromJSON("data\\copy_fsa.txt") %>% 
  as_tibble %>% 
  select(-geom) %>%
  filter(easting < 4e10) %>% # entry error: Camb - Mid Yell Voe - 2013-04-16 
  filter(easting > 0 & northing > 0) %>%
  filter(!is.na(date_collected)) %>%
  filter(karenia_mikimotoi >= 0) %>% # -99 in karenia...?
  mutate(date_collected=date(date_collected),
         year=year(date_collected),
         month=month(date_collected),
         yday=yday(date_collected),
         ydayCos=cos(yday/365),
         ydaySin=sin(yday/365),
         x_y=paste0(easting, "_", northing)) %>%
  filter(year < 2022)
fsa.sf <- st_as_sf(fsa.df, coords=c("easting", "northing"), crs=st_crs(27700))
# st_write(fsa.sf, "data\\fsa.shp", append=F)

fsa.ldf <- fsa.df %>% pivot_longer(8:12, names_to="spp", values_to="N")
```

```{r EDA_maps}
fsa.df %>% group_by(easting, northing) %>% 
  summarise(N=n(), nSin=n_distinct(sin), nSite=n_distinct(site), nArea=n_distinct(area)) %>%
  st_as_sf(coords=c("easting", "northing"), crs=st_crs(27700)) %>%
  # st_write("data\\fsa_xy_counts.shp", append=F)
  ggplot() + geom_sf(data=coast) +
  geom_sf(aes(colour=N)) +
  scale_colour_viridis_c()


geo_columns <- c("sin", "area", "site", "x_y")
locs <- map(geo_columns, ~unique(fsa.df[[.x]])) %>% setNames(geo_columns) 
map(locs, length)
ggplot(coast) + geom_sf() +
  geom_sf(data=filter(fsa.sf, sin %in% locs[[1]][1:10]), aes(colour=site))
ggplot(coast) + geom_sf() +
  geom_sf(data=filter(fsa.sf, area %in% locs[[2]][1:20]), aes(colour=area))
```



```{r EDA}
fsa.ldf %>% 
  group_by(year, month, spp) %>% 
  summarise(logN=mean(log(N+1), na.rm=T), 
            prP=mean(N>0, na.rm=T)) %>% 
  ggplot(aes(month, prP, group=year, colour=year)) + 
  stat_smooth(se=F) +
  facet_wrap(~spp) + 
  scale_colour_viridis_c() +
  scale_x_continuous(breaks=1:12)
fsa.ldf %>% 
  group_by(year, month, spp) %>% 
  summarise(logN=mean(log(N+1), na.rm=T), 
            prP=mean(N>0, na.rm=T)) %>% 
  ggplot(aes(month, prP, group=year, colour=year)) + 
  geom_line() +
  facet_wrap(~spp) + 
  scale_colour_viridis_c() +
  scale_x_continuous(breaks=1:12)
```

